\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-30 chapters/wcble.tex
Accuracy figure
Preface
Preface, ironed out overhangs

Template matching is generally preferred to peak picking.
The algorithm proposed by \textcite{Woody1967} in 1976 is a simple and elegant
latency estimation method that is still being used and is even
considered among the more performant techniques for
\acp{erp}~\cite{Thornton2008, Ouyang2017}.
It starts with an average \ac{erp} as a rough estimation of the aligned template
\ac{erp}, and iteratively refines this template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial \acp{erp} and aligning them
on these time points to form the next template.
Similarly, cross-correlation template matching can be combined with filtering.
\textcite{Souloumiac2013} uses XDAWN to construct a spatial filter before
template matching the spatially filtered signal. \textcite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{Gasser1983}.

@@

Some algorithms separate components into stimulus- and response-locked component clusters based on their
latency distributions~\cite{Jung2001, Takeda2008, Zhang1998, Yin2009}.
These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an \ac{erp} experiment, especially in
the case of \acp{bci}.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015, Wang2015, Ouyang2016, Ouyang2020}
is able to do this without entirely relying on the response time and can
separate stimulus-locked, response-locked, and multiple central component clusters.
Others, like spatial filtering methods that use decomposition methods to
determine the spatial filter, can be modified to separate
components~\cite{Ouyang2017}.

@@

Most of these methods suffer from a common drawback: they cannot be used in a
decoding scheme to improve performance for incoming epochs with unknown
class labels that contain either a target or non-target \ac{erp} response.
They can be used offline on a set of labeled epochs for testing hypotheses
concerning latency and jitter, for aligning templates, or for \ac{bci} performance
prediction, but not for \ac{erp} classification.
While some of the aforementioned methods could be adapted to perform
classification tasks, few studies investigate how to exploit this latency estimation
for jitter-resistant decoding.
\textcite{Hardiansyah2020} incorporated single-trial latencies in
classification by peak-picking within a given
\ac{erp} time window, unaware of the class of the epoch under investigation.
The \ac{cble} algorithm introduced
by \textcite{Thompson2012} also explicitly applies latency estimation in a
decoding setting.
\textcite{Thompson2012} initially formulated \ac{cble} as an offline performance
prediction method.
Later, its output was successfully adapted to compensate for jitter to improve
decoder performance~\cite{Mowla2017, Zisk2022}.

@@

\section{Methods}
\label{sec:wcble/methods}

@@

Consider a training set of $N$ EEG epochs with $C$ channels of $S$
samples $\{\mat{X}_n^\mathrm{train} \in \mathbb{R}^{C\times S}\}_{n=1}^N$
with corresponding training labels $\mathbf{l^\mathrm{train}} \in \{\mathrm{target},
	\textrm{non-target}\}_{n=1}^N$, and a
similar testing set of $M$ epochs $\{\mat{X}_m^\mathrm{test} \in
	\mathbb{R}^{C\times S}\}_{m=1}^M$.
We assume that the sampling period is $T$, i.e., that the sample with index $s$ is sampled at time $sT$.
In the following, we use the matrix slicing notation to denote row or column intervals extracted from a matrix.
For instance, $\mat{X}[:,s_1:s_2]$ denotes all columns of $\mat{X}$ with indices between $s_1$ (included) and $s_2$ (excluded).

@@

Instead, we opt for a more robust, probabilistic definition of class latencies.
This robust estimation method yields latencies that
\begin{enumerate*}[label=(\arabic*)]
  \item are more meaningful as input for the second-stage
    classifier, and
  \item lead to smoother convergence in our proposed iterative alignment scheme
    for \ac{wcble}, which heavily relies on exact latency estimation.
\end{enumerate*}
Assume classifier $\mathcal{C}(\theta,f,\Pr)$ now can also output a probability
per class $\Pr(l|
\mat{X}[:,s_1:s_2)], \theta)$ for a given epoch $\mat{X}$, a feature of many
common classifiers.
Analogous to equation~\ref{eq:score}, we can now write
\begin{equation}
\Pr(\mat{X},\theta,l,s) = \frac{1}{R}\Pr(\mat{X}[:,s:s+(s_2-s_1)],\theta,l)
  \quad \forall s\in 1,\ldots,R
	\label{eq:prob}
\end{equation}
The latency features assuming the epoch belongs to a given class
$l\in\{\textrm{target},\textrm{non-target}\}$ are then defined as the median of
the corresponding distributions
\begin{equation}
  s_l =\mathrm{median}\left[\Pr(s|\mat{X},\theta,l) \right]
  \label{eq:prob-latency}
\end{equation}
Note that $\Pr(s|\mat{X},\theta,\textrm{non-target}) =1-
  \Pr(s|\mat{X},\theta,\mathrm{target})$.
The median of the probability distribution over time is more robust to
outliers and noise than the maximum or minimum score.
For the non-target case, the median approach tends towards the center of a
near-uniform distribution, resulting in a more consistent latency estimate over
trials as compared to the minimum approach.

\subsection{\acl{wcble}}
\label{sec:wcble/methods/wcble}
To improve performance over \ac{cble}, we propose a new algorithm inspired both by
\ac{cble} and the aforementioned Woody iteration scheme termed \ac{wcble}.
Instead of using \ac{cble} to estimate the features of a second-stage classifier
directly, \ac{cble} latency estimation is used as a step in a Woody iteration scheme.
While the Woody algorithm iteratively enhances the \ac{snr} of an \ac{erp} template to
cross-correlate with the data, \ac{wcble} iteratively re-estimates the parameters of
the first-stage classifier.
To improve convergence and perform well in a classification setting, \ac{wcble}
aligns both targets and non-targets to their corresponding estimated latencies.

@@

\clearpage
\begin{subappendices}
\section{\Acs{cble} algorithm}
\label{sec:wcble/app/cble-alg}
\begin{algorithm}[h]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
  \mathbf{l^\mathrm{train}}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\State $\theta \gets
			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l^\mathrm{train}})$
      \algorithmiccomment{Train stage 1}
		\For{$n=1\ldots N$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\theta,
  \mat{Y}^\mathrm{train},
  \mathbf{s}_\mathrm{target}^\mathrm{train},
  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$

@@

\clearpage
\section{\Acs{wcble} algorithm}
\label{sec:wcble/app/wcble-alg}
\begin{algorithm}[h]
  \small
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l^\mathrm{train}},
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      \algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l^\mathrm{train}})$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor

>>>>

Template matching is generally preferred to peak picking.
The algorithm proposed by \textcite{Woody1967} in 1976 is a simple and elegant
latency estimation method that is still being used and is even
considered among the more performant techniques for
\acp{erp}~\cite{Thornton2008, Ouyang2017}.
It starts with an average \ac{erp} as a rough estimation of the aligned template
\ac{erp}, and iteratively refines this template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial \acp{erp} and aligning them
on these time points to form the next template.
Similarly, cross-correlation template matching can be combined with filtering.
\textcite{Souloumiac2013} use the XDAWN technique to construct a spatial filter before
template matching the spatially filtered signal. \textcite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{Gasser1983}.

@@

Some algorithms separate components into (sets of) stimulus- and response-locked component clusters based on their
latency distributions~\cite{Jung2001, Takeda2008, Zhang1998, Yin2009}.
These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an \ac{erp} experiment, especially in
the case of \acp{bci}.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015, Wang2015, Ouyang2016, Ouyang2020}
is able to do this without entirely relying on the response time and can
separate stimulus-locked, response-locked, and multiple central component clusters.
Others, like spatial filtering methods that use decomposition methods to
determine the spatial filter, can be modified to separate
components~\cite{Ouyang2017}.

@@

Most of these methods suffer from a common drawback: they cannot be used in a
decoding scheme to improve performance for incoming epochs with unknown
class labels that contain either a target or non-target \ac{erp} response.
They can be used offline on a set of labeled epochs for testing hypotheses
concerning latency and jitter, for aligning templates, or for \ac{bci} performance
prediction, but not for \ac{erp} classification.
While some of the aforementioned methods could be adapted to perform
classification tasks, few studies investigate how to exploit this latency estimation
for jitter-resistant decoding.
\textcite{Hardiansyah2020} incorporated single-trial latencies in
classification by peak-picking within a given
\ac{erp} time window, unaware of the class of the epoch under investigation.
The \ac{cble} algorithm by \textcite{Thompson2012} also explicitly applies latency estimation in a
decoding setting.
\textcite{Thompson2012} initially formulated \ac{cble} as an offline performance
prediction method.
Later, its output was successfully adapted to compensate for jitter to improve
decoder performance~\cite{Mowla2017, Zisk2022}.

@@

\section{Materials \& methods}
\label{sec:wcble/methods}

@@

Consider a training set of $N$ EEG epochs with $C$ channels of $S$
samples $\{\mat{X}_n^\mathrm{train} \in \mathbb{R}^{C\times S}\}_{n=1}^N$
This dataset has a corresponding set of training labels $\mathbf{l^\mathrm{train}} \in \{\mathrm{target},
	\textrm{non-target}\}_{n=1}^N$, and a
similar testing set of $M$ epochs $\{\mat{X}_m^\mathrm{test} \in
	\mathbb{R}^{C\times S}\}_{m=1}^M$.
We assume that the sampling period is $T$, i.e., that the sample with index $s$ is sampled at time $sT$.
In the following, we use the matrix slicing notation to denote row or column intervals extracted from a matrix.
For instance, $\mat{X}[:,s_1:s_2]$ denotes all columns of $\mat{X}$ with indices between $s_1$ (included) and $s_2$ (excluded).

@@

Instead, we opt for a more robust, probabilistic definition of class latencies.
This robust estimation method yields latencies that
\begin{enumerate*}[label=(\arabic*)]
  \item are more meaningful as input for the second-stage
    classifier, and
  \item lead to smoother convergence in our proposed iterative alignment scheme
    for \ac{wcble}, which heavily relies on exact latency estimation.
\end{enumerate*}
Assume classifier $\mathcal{C}(\theta,f,\Pr)$ now can also output a probability
per class $\Pr(l|
\mat{X}[:,s_1:s_2)], \theta)$ for a given epoch $\mat{X}$, a feature of many
common classifiers.
Analogous to equation~\ref{eq:score}, we can now write
\begin{equation}
\Pr(\mat{X},\theta,l,s) = \frac{1}{R}\Pr(\mat{X}[:,s:s+(s_2-s_1)],\theta,l)
  \quad \forall s\in 1,\ldots,R
	\label{eq:prob}
\end{equation}
The latency features assuming the epoch belongs to a given class
$l\in\{\textrm{target}, \textrm{non-target}\}$ are then defined as the median of
the corresponding distributions
\begin{equation}
  s_l =\mathrm{median}\left[\Pr(s|\mat{X},\theta,l) \right]
  \label{eq:prob-latency}
\end{equation}
Note that $\Pr(s|\mat{X},\theta,\textrm{non-target}) =1-
  \Pr(s|\mat{X},\theta,\mathrm{target})$.
The median of the probability distribution over time is more robust to
outliers and noise than the maximum or minimum score.
For the non-target case, the median approach tends towards the center of a
near-uniform distribution, resulting in a more consistent latency estimate over
trials as compared to the minimum approach.

\subsection{\acl{wcble}}
\label{sec:wcble/methods/wcble}
To improve performance over \ac{cble}, we propose \ac{wcble} a new algorithm inspired both by
\ac{cble} and the aforementioned Woody iteration scheme.
Instead of using \ac{cble} to estimate the features of a second-stage classifier
directly, \ac{cble} latency estimation is used as a step in a Woody iteration scheme.
While the Woody algorithm iteratively enhances the \ac{snr} of an \ac{erp} template to
cross-correlate with the data, \ac{wcble} iteratively re-estimates the parameters of
the first-stage classifier.
To improve convergence and perform well in a classification setting, \ac{wcble}
aligns both targets and non-targets to their corresponding estimated latencies.

@@

\clearpage
\begin{subappendices}
\section{\Acs{cble} algorithm}
\label{sec:wcble/app/cble-alg}
\begin{algorithm}[h]
  \sffamily
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
  \mathbf{l^\mathrm{train}}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\State $\theta \gets
			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l^\mathrm{train}})$
      \algorithmiccomment{Train stage 1}
		\For{$n=1\ldots N$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\theta,
  \mat{Y}^\mathrm{train},
  \mathbf{s}_\mathrm{target}^\mathrm{train},
  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$

@@

\clearpage
\section{\Acs{wcble} algorithm}
\label{sec:wcble/app/wcble-alg}
\begin{algorithm}[h]
  \sffamily
  \small
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l^\mathrm{train}},
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      \algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l^\mathrm{train}})$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor
